```{r setup, include=FALSE}
require(ggplot2)
require(readr)

theme_set(theme_grey(base_size=20))

# Covariance matrix for noise "like" the given residuals.
#
# Assumes they were generated from a stationary process.
EmpiricalCovarianceMatrix <- function(residuals) {
  autocorrelation <- acf(residuals, lag.max=length(residuals))$acf
  sigma_sq <- var(residuals)
  return (outer(1:length(residuals), 1:length(residuals),
                function(a, b) { sigma_sq * autocorrelation[abs(a - b) + 1] }))
}

# A matrix whose columns are the values of the supplied basis functions,
# evaluated at the supplied points x.
LinearLeastSquaresMatrix <- function(x, functions) {
  sapply(functions,
         function(f) {
           sapply(x, function(x) { f(x) })
         })
}

# A matrix for the disconnected linear model.
DisconnectedLinearMatrix <- function(x, x_break) {
  functions <- list(function(z) 1 * (z < x_break),
                    function(z) as.numeric(z) * (z < x_break),
                    function(z) 1 * (z > x_break),
                    function(z) as.numeric(z) * (z > x_break))
  LinearLeastSquaresMatrix(x, functions)
}

# A matrix for the disconnected linear model.
ConnectedLinearMatrix <- function(x, x_break) {
  functions <- list(function(z) 1,
                    function(z) z,
                    function(z) as.numeric(z - x_break) * (z > x_break))
  LinearLeastSquaresMatrix(x, functions)
}

# A matrix for a single trendline model.
SingleLinearMatrix <- function(x) {
  functions <- list(function(z) 1,
                    function(z) z)
  LinearLeastSquaresMatrix(x, functions)
}

# The "best fit" value (for the given model) for each y value.
#
# Args:
#   y:  A vector of datapoints.
#   model:  A model matrix: each column gives the value of one of the basis
#     functions, evaluated at the same position as the corresponding y value.
Fit <- function(y, model) {
  model %*% solve(t(model) %*% model) %*% t(model) %*% y
}

# Build a data.frame from an RSS CSV data file.
#
# Args:
#   rss_file_name:  The name of the RSS CSV file (columns 'year', 'month',
#     'anomaly').
#   cutoff_date:  The date when the rate of warming is supposed to have
#     changed.
RssDataset <- function(rss_file_name, cutoff_date) {
  # Read the filename.
  d <- read_csv(rss_file_name)
  # Compute a proper 'date' column.
  d$date <- as.Date(sprintf("%04d-%02d-01", d$year, d$month))

  # Compute the fit points for a "single trendline" model.
  d$single <- Fit(d$anomaly,
                  SingleLinearMatrix(d$date))

  # Compute the fit points for the "disconnected trendlines" model.
  d$disconnected <- Fit(d$anomaly,
                        DisconnectedLinearMatrix(d$date, cutoff_date))

  # Compute the fit points for the "connected trendlines" model.
  d$connected <- Fit(d$anomaly,
                     ConnectedLinearMatrix(d$date, cutoff_date))

  return (d)
}

# The number of degrees per century between the earliest and latest dates which
# pass the filter.
#
# Args:
#   dates:  A vector of dates.
#   y:  A vector of values (presumably temperatures or temperature anomalies),
#     one for each value in dates.
#   filter:  A boolean function which accepts a date.
DegreesPerCenturyWithFilter <- function(dates, y, filter) {
  i_ok <- which(filter(dates))
  i_extreme <- which(dates %in% range(dates[i_ok]))
  days_per_century <- 36524.25
  diff_days <- diff(as.numeric(dates[i_extreme]))
  return (diff(y[i_extreme]) / (diff_days / days_per_century)) 
}

# The number of degrees per century between the earliest and latest dates.
#
# Args:
#   dates:  A vector of dates.
#   y:  A vector of values (presumably temperatures or temperature anomalies),
#     one for each value in dates.
DegreesPerCentury <- function(dates, y) {
  DegreesPerCenturyWithFilter(dates, y, function(d) as.logical(d) | TRUE)
}

# How many years of warming (at a given rate) to get a change in temperature?
YearsOfWarming <- function(temp_change, rate_per_century) {
  temp_change / rate_per_century * 100
}

# Append a 'degrees C' to a temperature
PrintTemp <- function(temperature) {
  sprintf('%4.2f°C', temperature)
}

# Print the trend in human-readable form: round to 2 decimal places, and label
# as degrees per century.
PrintTrend <- function(temperature, time_span='century') {
  sprintf('%s / %s', PrintTemp(temperature), time_span)
}

# A geom_text to label a line.
TextLabel <- function(label, d, column, ...) {
  geom_text(label=label,
            data=d[round((1 + nrow(d)) / 2), ],
            aes_string(x='date', y=column),
            hjust=0, vjust=1.5, size=10, ...)
}

# A geom_line to show a fit.
FitLine <- function(d, column, ...) {
  geom_line(data=d, aes_string(x='date', y=column), size=1.2, ...)
}

# The (first) index of the element in y whose following element increases by
# the most.
BiggestJumpIndex <- function(y) {
  dy <- diff(y)
  head(which(dy == max(dy)), 1)
}

# Cutoff date variable, and associated functions.
cutoff_date <- as.Date("1997-02-15")
Before <- function(d) d < cutoff_date
After <- function(d) d > cutoff_date

# Construct the RSS data frame.
rss_file <- 'content/blog/connect_your_trendlines/data/rss.csv'
d_rss <- RssDataset(rss_file, cutoff_date)

single_trendline_slope <- DegreesPerCentury(d_rss$date, d_rss$single)
bad_slope_before <- DegreesPerCenturyWithFilter(d_rss$date, d_rss$disconnected,
                                                Before)
bad_slope_after <- DegreesPerCenturyWithFilter(d_rss$date, d_rss$disconnected,
                                               After)
good_slope_before <- DegreesPerCenturyWithFilter(d_rss$date, d_rss$connected,
                                                 Before)
good_slope_after <- DegreesPerCenturyWithFilter(d_rss$date, d_rss$connected,
                                                After)
jump <- max(diff(d_rss$disconnected))
```

- Stuff to cover in the intro...
  - "No warming for 18 years!" claims.

Lately, various scientists have argued that the "pause" was an illusion.
They've used a variety of techniques (TODO: list the main ones).

As a layperson, I find these esoteric corrections hard to evaluate.
So let's put them aside for this post, and ask a simpler question.

*From a "pure" data analysis perspective, did even the most "pause-friendly" dataset ever show a pause in the first place?*

The answer is a surprisingly firm "no".

This seems absurd: *if* it didn't, then *why* did people think global warming had stopped in the first place?
We will see that this claim was based on a data analysis error -- a subtle one, but easily understood (and, ultimately, inarguable).

## The Data: RSS

To begin, here are the data.

```{r bare_rss, echo=FALSE}
p <- (ggplot(data=d_rss, aes(x=date))
      + geom_line(colour='#888888', aes(y=anomaly))
      + ggtitle('RSS Satellite Temperature Record')
      + scale_x_date('Date')
      + scale_y_continuous('Temperature Anomaly (K)')
      )
print(p)
```

There's a lot of fluctuation here, due to ((give reasons.  Natural variability?  Volcanoes?  El Nino?)).
It looks like the later data are probably higher than the earlier data, but it's hard to be sure.

To figure out what's going on "underneath" these fluctuations, we need to treat them as noise and look for a buried signal.

We need *models*.

## The simplest model: one trendline

One approach is to compute the line of best fit.
Basically, we're *assuming* the trend has been constant over time, and asking which trend best describes that data.

```{r single_trendline, echo=FALSE}
colour_single <- 'purple'
column_single <- 'single'
line_single <- FitLine(d=d_rss, column=column_single, colour=colour_single)
label_single <- TextLabel(PrintTrend(single_trendline_slope), d_rss, 'single',
                          colour=colour_single)
print(p
      + line_single
      + label_single
      + ggtitle('RSS fit: single trendline')
      )
```

Overall, the RSS data show a slope of `r PrintTrend(single_trendline_slope)`.

This is an exceedingly simple model -- and that is its strength!
More complicated models risk mistaking noise for signal.
This phenomenon is called "overfitting", and we want to avoid it.

However, we also need to be avoid using a model that's *too* simple.

## The next model: the trend changes

Suppose the underlying trend *changes*.
Our model will still give an answer, but it will be wrong.
We need a more sophisticated model to check whether the trend has changed.

(Note that it is very important to be able to detect trendline changes.
After all, changing the trend is *exactly what environmentalists are trying to do!*)

In particular, [Lord Monckton suspects](http://www.climatedepot.com/2015/11/04/no-global-warming-at-all-for-18-years-9-months-a-new-record-the-pause-lengthens-again-just-in-time-for-un-summit-in-paris/) that the trend has completely flattened (i.e., global warming has stopped).
He looks for the longest stretch of time which yields a flat trend up to the present day.
By the same method we used above, he obtains a shockingly long answer: he finds *no global warming for 18 years!*

```{r partial_disconnected, echo=FALSE}
colour_disc <- 'blue'
column_disc <- 'disconnected'
line_disc <- FitLine(d=d_rss, column=column_disc, colour=colour_disc)
label_disc <- TextLabel(PrintTrend(bad_slope_after), d_rss, column_disc,
                        colour=colour_disc)
p_disconnected <- (p
                   + line_disc
                   + label_disc
                   + ggtitle('RSS fit (latest 18 years only):\n"No" Global Warming!')
                   )
print(p_disconnected + coord_cartesian(xlim=c(cutoff_date + 30, max(d_rss$date))))
```


Actually, it's not *exactly* the same method; he has excluded the earlier dates.
When we add them back in, we find a fatal flaw in his approach.

```{r full_disconnected, echo=FALSE}
i_jump <- BiggestJumpIndex(d_rss$disconnected)
column_jump <- column_disc
colour_jump <- 'red'
print(p_disconnected
      + ggtitle('Fitting the old data too:\nInstantaneous temperature jump')
      + TextLabel(PrintTrend(bad_slope_before), d_rss[1, ], column_disc,
                  colour=colour_disc)
      + FitLine(d=d_rss[i_jump + 0:1, ], column=column_jump,
                colour=colour_jump)
      + TextLabel(sprintf('+%s!', PrintTemp(jump)), d_rss[i_jump, ],
                  column_jump, colour=colour_jump)
      )
```


We do see that the slope has changed (from `r PrintTemp(bad_slope_before)` to `r PrintTrend(bad_slope_after)`).
But we see something else: the temperature *instantaneously jumps* by `r PrintTemp(jump)` at the transition!

## Fixing the model: trendlines done *right*

Everybody agrees that the *real* temperature trend doesn't have any "jumps".
So why not bake this assumption into the model?

```{r connected, echo=FALSE}
colour_conn <- 'red'
column_conn <- 'connected'
line_conn <- FitLine(d=d_rss, column=column_conn, colour=colour_conn)
label_conn_before <- TextLabel(PrintTrend(good_slope_before),
                               d_rss[nrow(d_rss)*0.15, ], column_conn,
                               colour=colour_conn)
label_conn_after <- TextLabel(PrintTrend(good_slope_after),
                              d_rss[nrow(d_rss)*0.65, ], column_conn,
                              colour=colour_conn)

p_connected <- (p
                + line_conn
                + label_conn_before
                + label_conn_after
                + ggtitle('"Changing trendlines" model:\nthe correct (i.e., connected) version')
                )
print(p_connected)
```

The *improved* model still allows the trend to change; it just makes sure to *connect the trendlines*.
We see that the warming is alive and well: `r PrintTrend(good_slope_after)`.
This is somewhat smaller than the average slope from the simplest model (`r PrintTrend(single_trendline_slope)`)

However, it's actually *bigger* than the biggest slope from the disconnected model (`r PrintTrend(bad_slope_before)`)!
This is another effect of the disconnected model squeezing all the warming into a single instant.
It underestimates the warming *before* as well as after.

## Let's plot all the models together

```{r all_models, echo=FALSE}
print(p
      + line_single
      + line_disc
      + line_conn
      + ggtitle('All models together')
      )
```

Overall, the "connected trendline" model looks really similar to the single-trendline model.

By contrast, the disconnected trendline model looks downright bizarre.
Of course, it doesn't need the other models to make it look weird!
All it needed was for the implicit "jump" to be exposed: anyone can see at a glance that it's wrong.

## What's the longest pause with the *fixed* model?

We can play the same game as Lord Monckton, but using the better model.
How far back can we go and find a zero-slope trendline *which connects to the previous trendline*?


```{r pause_hunting, include=FALSE, cache=TRUE}
last_date <- tail(d_rss$date, 1)
seen_negative_slope <- FALSE
for (pause_date in seq(from=last_date - 45, to=head(d_rss$date, 1),
                       by='-1 month')) {
  connected_fit <- Fit(d_rss$anomaly,
                       ConnectedLinearMatrix(d_rss$date, pause_date))
  post_cutoff_slope <- DegreesPerCenturyWithFilter(d_rss$date, connected_fit,
                                                   function(d) d > pause_date)
  if (seen_negative_slope && post_cutoff_slope > 0) {
    break
  }
  seen_negative_slope <- seen_negative_slope || (post_cutoff_slope < 0)
}

# It is annoying that I have to do this.  Why doesn't a sequence of Date
# objects automatically contain Date objects?
pause_date <- as.Date(pause_date, origin='1970-01-01')

# Compute the fit and slopes for this pause.
d_rss$newpause <- Fit(d_rss$anomaly,
                      ConnectedLinearMatrix(d_rss$date, pause_date))
pause_slope_before <- DegreesPerCenturyWithFilter(d_rss$date, d_rss$newpause,
                                                  function(d) d < pause_date)
pause_slope_after <- DegreesPerCenturyWithFilter(d_rss$date, d_rss$newpause,
                                                  function(d) d > pause_date)

# Quick-and-dirty utility functions to help answer the question, "how long, in
# months, from one date to another?" where we only look at the year and month
# of each date.
Month <- function(date) as.numeric(format(date, '%m'))
Year <- function(date) as.numeric(format(date, '%Y'))
MonthsSpan <- function(from, to) {
  12 * (Year(to) - Year(from)) + Month(to) - Month(from)
}
PrintMonthsSpanInYears <- function(from, to) {
  months <- MonthsSpan(from, to)
  sprintf('%d years and %d months', months %/% 12, months %% 12)
}
```

```{r connected_pause, echo=FALSE}
colour_newpause <- 'orange'
column_newpause <- 'newpause'
line_newpause <- FitLine(d=d_rss, column=column_newpause,
                         colour=colour_newpause)
label_newpause_before <- TextLabel(PrintTrend(pause_slope_before),
                               d_rss[nrow(d_rss)*0.15, ], column_newpause,
                               colour=colour_newpause)
label_newpause_after <- TextLabel(PrintTrend(pause_slope_after),
                              d_rss[nrow(d_rss)*0.65, ], column_newpause,
                              colour=colour_newpause)
print(p
      + line_newpause
      + label_newpause_before
      + label_newpause_after
      + ggtitle(sprintf("The longest 'pause' with a sensible model:\n%s",
                        PrintMonthsSpanInYears(pause_date, last_date)))
      )
```

This corrected "pause" extends back to
`r format(pause_date + 30, '%B %Y')`, a span of
`r PrintMonthsSpanInYears(pause_date, last_date)`.

## Other questions worth investigating

Here are some related questions which I haven't evaluated:

- Has the rate of global warming *slowed?*
- Have the IPCC climate models overestimated the recent global warming?

These questions are interesting, important, and well worth pursuing!
But this blog post is tightly focused on one question only:

## Did global warming **stop** for 18 years?

No. 

The RSS satellite record unambiguously shows global warming, including to the present day.
As I've shown, the idea that we had "no warming for 18 years" actually implies that 18 years' worth of warming happened instantaneously.

Squeeze the warming one place (the slope), and it pops up somewhere else (the "jump").

If anyone wants to find an *actual* pause, they will have to use connected trendlines.
