```{r setup, include=FALSE}
require(knitr)
opts_chunk$set(tidy=FALSE)
# Don't use scientific notation.  For this post, we'd rather use integers.
options('scipen'=20)
```

Recently, a friend asked me a fun math question. He's creating a real-life RPG
for artists, called _illolife_[^1].  He wrote the experience point threshold for
each level in terms of the previous level's threshold.  His question: is there a
formula to get that threshold _directly_, without computing all the levels
before it?

I say it's "fun" because the level differences follow the [Fibonacci sequence].
Specifically: If $X_i$ represents the experience needed for level $i$[^2], and
$F_i$ is the $i$th Fibonacci number, then
$$X_{i + 1} = X_i + 250 F_i.$$
(The factor of 250 is just a scaling factor, and it's not important.)

So the question boils down to: _is there a compact way to compute sums of
Fibonacci numbers_?

## Fibonacci sums are _almost_ Fibonacci numbers

As it happens, Fibonacci _sums_ follow the same pattern as Fibonacci _numbers_,
but they're always ahead 2 spots in the sequence, and off by 1:
$$\sum_{i=1}^n F_i = F_{n + 2} - 1.$$
Here are a few examples.

+-----+-------------------------------+-------------+
| $n$ | $\sum\limits_{i=1}^n F_i$     | $F_{n + 2}$ |
+=====+===============================+=============+
| $1$ | $\mathbf{1}$                  | $2$         |
+-----+-------------------------------+-------------+
| $2$ | $1 + 1 = \mathbf{2}$          | $3$         |
+-----+-------------------------------+-------------+
| $4$ | $1 + 1 + 2 + 3 = \mathbf{7}$  | $8$         |
+-----+-------------------------------+-------------+
| $8$ | $1 + 1 + 2 + 3 + 5            | $55$        |
|     |  + 8 + 13 + 21 = \mathbf{54}$ |             |
+-----+-------------------------------+-------------+

What a delightful result!  It shows that Fibonacci sums are as easy to compute
as Fibonacci numbers.  This raises the question -- how easy are Fibonacci
numbers to compute?

## Binet's formula and the golden ratio

A simple expression, known as *[Binet's formula]*, can directly compute the $n$th
Fibonacci number.  Using $\varphi \equiv \left(1 + \sqrt{5}\right) / 2$, the
[golden ratio], we have:

$$F_n = \frac{\varphi ^ n - \left(1 - \varphi\right) ^ n}{\sqrt{5}}.$$

Here is a simple [R] function to compute it.

```{r fib_closed_form}
fib_closed_form <- function(n) {
  # Compute the nth fibonacci number with Binet's formula, using the golden
  # ratio 'phi'.
  phi <- (1 + sqrt(5)) / 2
  return ((phi ^ n - (1 - phi) ^ n) / sqrt(5))
}
```

Binet's formula surprised me at first: it's a function of _irrational_ numbers
which gives _integers_ every time!  Fortunately, it's fairly easy to work them
out by hand to see why it works.  I did the first few, which made it seem a lot
more plausible to me.

I was worried that we might lose precision by going to floating point numbers
and then coming back to integers.  To check whether this is a problem, we need
a "gold standard" for Fibonacci numbers to compare against.  Here is another
helper function which computes Fibonacci numbers directly from their definition.
(Since it only uses integers, it shouldn't be affected by floating point
errors.)

```{r fib_direct}
fib_direct <- function(n) {
  # Compute the nth fibonacci number directly from the definition.
  # (Assumes n >= 1.)
  fib <- c(1, 1)
  while (length(fib) < max(n)) {
    fib <- c(fib, sum(tail(fib, 2)))
  }
  return (fib[n])
}
```

Now we can compare the results from each.

```{r fib_compare}
# A set of n-values for testing purposes.
n <- seq(from=5, to=80, by=5)

# Compute Fibonacci numbers at these n-values by both methods, and compute the
# difference between them.
fib_compare <- data.frame(n=n,
                          direct=fib_direct(n),
                          binet=fib_closed_form(n))
fib_compare$diff <- fib_compare$binet - fib_compare$direct 

# Display the results.
print(fib_compare, row.names=FALSE)
```

Looks like the concern was justified: things go awry somewhere between $F_{70}$
and $F_{75}$.  Still, the formula works remarkably well until that point.
Especially considering the application -- it'll be a long time before any
illolifer hits level 70!

## A surprising connection

So far, we've used two convenient results:

1. The sums of Fibonacci numbers are (almost) Fibonacci again.
2. Binet's formula, $F_n = \left(\varphi^n - (1 - \varphi)^n\right) / \sqrt{5}$,
   directly gives the $n$th Fibonacci number.

Surprisingly, these results are related!

Consider the first one: the sum over a sequence gives the same sequence[^3].
For _continuous_ quantities (instead of _discrete_), we'd have integrals instead
of sums, and functions instead of sequences.  So, this is like saying the
_integral_ of a _function_ gives that same _function_.  This is a well-known
property of [exponential functions]!  How suggestive -- do the Fibonacci numbers
grow exponentially?

They do.  Look at Binet's formula, and realize that
$\left|1 - \varphi\right| < 1$; this means that the
$\left(1 - \varphi\right)^n$ term is insignificant for large $n$.  What remains
is
$$F_n \approx \frac{\varphi^n}{\sqrt{5}} \,\,\,\,\,\,\,\,(\text{for large }n).$$
This is exponential with a base of $\varphi$.

So we needed two properties of Fibonacci numbers, and it turns out that their
hidden exponential nature underpins both of these properties.

## Finally: computing the experience for level $n$

Putting all the pieces together, and remembering that
$\varphi = \left(1 + \sqrt{5}\right) / 2$, we have

$$X_n = 250\left(
  \frac{\varphi^{n + 1} - \left(1 - \varphi\right)^{n + 1}}{\sqrt{5}} - 1
\right)$$

[^1]: This is [the link for
illolife](https://www.facebook.com/groups/illolifeRPG/#_=_), but note that you
have to be logged in to facebook to see anything.
[^2]: $X_1 = 0$, since everybody starts at level 1.
[^3]: Well, up to a small constant offset.

Sanity check: what are the first few levels?

```{r}
X <- function(n) {
  phi <- (1 + sqrt(5)) / 2
  return (250 * ((phi ^ (n + 1) - (1 - phi) ^ (n + 1)) / sqrt(5) - 1))
}
cat(X(1:10))
```

These numbers exactly match the ones he provided me.  Success!

[Fibonacci sequence]: http://en.wikipedia.org/wiki/Fibonacci_number
[exponential functions]: http://www.animations.physics.unsw.edu.au/jw/calculus.htm#exp
[golden ratio]: http://www.mathsisfun.com/numbers/golden-ratio.html
[R]: http://www.r-project.org/
[Binet's formula]: http://mathworld.wolfram.com/BinetsFibonacciNumberFormula.html
