I recently found an interesting graph comparing countries' understanding of evolution with their GDP.
More GDP _should_ mean better education, which _should_ mean more people who understand basic scientific concepts (like evolution).
That's more or less what the graph shows, but there's a huge outlier.

<blockquote class="twitter-tweet" lang="en"><p>There is a close relationship between GDP and an understanding of the theory of evolution – with one exception. <a href="https://t.co/stD3lBfqd1">pic.twitter.com/stD3lBfqd1</a></p>&mdash; Max Roser (@MaxCRoser) <a href="https://twitter.com/MaxCRoser/status/589845245862477824">April 19, 2015</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

The USA massively bucks the trend: despite boasting the second-highest GDP, we have the second-*lowest* rate of accepting evolution.
Does the United States truly stand out as much as it appears?

## A few missing datapoints

In this dataset, the USA isn't just an evolutionary outlier; it's a geographical one, too.
32 of the 34 countries are European: all except for Japan and the USA.
The omission of Canada, which is similar both culturally and geographically, is especially startling.

This suspiciously skewed distribution raises a question: has the author _manufactured_ an outlier by leaving out all the similar countries?
The answer is no, for at least two reasons.

First, consider the source.
The data come from [a 2006 Science paper](https://www.sciencemag.org/content/313/5788/765.summary) which combined a variety of surveys from recent years.
The 32 European countries come from the [Eurobarometer survey 63.1](https://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/4563).
The remaining two countries correspond to the nationalities of the authors (two American and one Japanese).
Thus, it seems more likely that the US and Japan were added to the diverse Eurobarometer data, rather than any countries being omitted.

Second, even if we do include Canada, it doesn't make the US any less of an outlier.
[Canadian belief in evolution](http://angusreid.org/britons-and-canadians-more-likely-to-endorse-evolution-than-americans/) is significantly higher than in the US (61% vs. 40%), while their [GDP](https://www.cia.gov/library/publications/the-world-factbook/rankorder/2004rank.html) is somewhat lower (\$44,500 vs. \$54,800).
Canada is well in line with the rest of the data.

```{r setup, include=FALSE}
require(ggplot2)
require(reshape2)
require(scales)

theme_set(theme_grey(base_size=20))

# Estimate the index of x within values, returning fractional parts if it's
# closer to one side.
#
# Assumes the values are sorted.
#
# Args:
#   x:  A number.
#   values:  A sorted numeric vector.
FractionalIndex <- function(x, values) {
  if (x < values[1]) {
    return (NA)
  }
  for (i in 2:length(values)) {
    if (x <= values[i]) {
      lower <- values[i - 1]
      upper <- values[i]
      return (i - 1 + (x - lower) / (upper - lower))
    }
  }
  return (NA)
}

# Convert the coordinates of a point to GDP and evolution belief figures.
#
# Args:
#   d:  A data.frame with (at least) the following columns.
#     x:  A number representing the x-coordinate of the point.
#     y:  A number representing the y-coordinate of the point.
#
# Returns:
#   The GDP of a country plotted at x.
ComputeGdpAndEvo <- function(d) {
  d$GDP <- d$x
  d$evo <- d$y
  for (i in 1:nrow(d)) {
    d$GDP[i] <- 5000 * (1 + FractionalIndex(d$x[i], c(98.92, 150.26, 202.51,
                                                      254.32, 307.64, 358.60,
                                                      414.08, 464.38, 515.98,
                                                      568.23)))
    d$evo[i] <- 10 * (1 + FractionalIndex(d$y[i], c(108.38, 155.87, 204.45,
                                                    253.24, 301.82, 348.66,
                                                    398.53, 446.46)))
  }
  return (d[, !(names(d) %in% c('x', 'y'))])
}

# Indices of points which are "extreme" in one or more of these dimensions.
#
# Args:
#   data:  A data.frame.
#   column_names:  The names of the columns to check for outliers.
#   num_sd:  The number of standard deviations to be considered an 'outlier'.
#
# Returns:
#   The indices of the extreme points.
Outliers <- function(data, column_names, num_sd=1) {
  i <- c()
  for (column_name in column_names) {
    column <- data[[column_name]]
    bounds <- median(column) + num_sd * sd(column) * c(-1, 1)
    i <- union(i, which(column < bounds[1] | column > bounds[2]))
  }
  return (i)
}

file <- 'content/blog/evolution_vs_education/data/max_roser_evolution_gdp.txt'
d <- ComputeGdpAndEvo(read.table(file, sep='\t', header=TRUE))
```

```{r gdp_evo, warning=FALSE, echo=FALSE}
d.canada <- rbind(d, data.frame(Country='Canada', GDP=44500, evo=61,
                                Reading=NA, Mathematics=NA, Science=NA))
i <- c(Outliers(d.canada, c('GDP', 'evo'), num_sd=1.5),
       which(d.canada$Country == 'Canada'))
p <- (ggplot(data=d.canada, aes(x=GDP, y=evo))
      + geom_point(size=3)
      + geom_text(data=d.canada[i, ], aes(label=Country), vjust=1.3)
      + scale_x_continuous('Gross Domestic Product per Capita', labels=dollar)
      + scale_y_continuous('Percent of the Public that Believes in Evolution')
      + ggtitle('Belief in Evolution versus National Wealth\n(Canada too!)')
      )
print(p)
```

Of course, there are many other countries in the world. 
It would be fascinating to get good survey data for them, to see whether the US is as exceptional as it seems.
But among the countries we do have, the United States really does seem to stand out.

## Are we just measuring education?

A friend suggested to me that maybe this has nothing to do with evolution in particular.
Maybe this evolution score merely reflects lousy education in general.
An intriguing suggestion!

To find out, I checked the 2010 OECD World Education Rankings, which [The Guardian visualizes nicely here](https://www.theguardian.com/news/datablog/2010/dec/07/world-education-rankings-maths-science-reading).
The US indeed scores poorly for Mathematics, but puts up respectable numbers for Reading and Science. 
The [full dataset](https://docs.google.com/spreadsheet/ccc?key=0AonYZs4MzlZbdEMzTjN5cHY1MmlJOHI3cmZCamRQWEE&hl=en#gid=1) contains 65 countries, including all but two from the evolution graph[^missing].
Can any of these explain the USA's outlier position?

```{r evo_edu, fig.height=15, warning=FALSE, echo=FALSE, include=FALSE}
d <- d[complete.cases(d), ]
d.melt <- melt(d, id.vars=c('Country', 'GDP', 'evo'))
i.out <- Outliers(d, c('GDP', 'evo', 'Reading', 'Science', 'Mathematics'),
                  num_sd=1.5)
outlier.countries <- d$Country[i.out]
p <- (ggplot(data=d.melt, aes(x=value, y=evo))
      + geom_point()
      + geom_text(data=d.melt[which(d.melt$Country %in% outlier.countries), ],
                  aes(hjust=ifelse(Country == 'Bulgaria', -0.1, 0.5),
                      label=Country),
                  vjust=1.3)
      + facet_wrap(~ variable, ncol=1)
      + scale_x_continuous('2010 OECD Score')
      + scale_y_continuous('Belief in Evolution')
      + ggtitle('Belief in Evolution vs. 2010 OECD Scores')
      )
print(p)
```

```{r gdp_edu, fig.height=15, warning=FALSE, echo=FALSE}
p <- (ggplot(data=d.melt, aes(x=GDP, y=value))
      + geom_point()
      + geom_text(data=d.melt[which(d.melt$Country %in% outlier.countries), ],
                  aes(hjust=ifelse(Country == 'Bulgaria', -0.1, 0.5),
                      label=Country),
                  vjust=1.3)
      + facet_wrap(~ variable, ncol=1)
      + scale_x_continuous(labels=dollar)
      + scale_y_continuous('2010 OECD Score')
      + ggtitle('2010 OECD Scores vs. GDP')
      )
print(p)
```

While the US is a little low in Mathematics (middle graph), it's certainly not an outlier.
And the other two scores -- Reading and Science -- are precisely in line with the rest of the data.

No matter how you slice it, America's dismal understanding of evolution can't be explained by its solidly mediocre education levels.

[^missing]: Sorry, Cyprus and Malta.
